Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-19 papers/wdmerger_I/paper.tex
In the paper directory, keep a notes directory with LaTeX source of any notes that are supplementary to the main paper.
Update the free fall figure.

A simple dynamical test to verify the gravitation physics in CASTRO is
the case of gravitational free fall. We place two stars on the grid 
in the manner of Section \ref{sec:initial_models}. The distance $a$ between 
them corresponds to a chosen orbital period $T$, consistent with the total
system mass $M$, but we disable the rotational source terms so that 
the stars start at rest in an inertial reference frame. 
Thus the stars will simply begin moving toward each other.
As long as the stars remain approximately spherical, the stars can be 
treated as point masses (this only seriously breaks down after the stars
have come into contact). In dimensionless units where $r \to r / a$ and 
$t \to 2\sqrt{2}\pi t / T$, the simple free fall equation of motion governing the
distance $r$ between their centers of mass takes the form:
\begin{equation}
  \ddot{r}(t) = - \frac{1}{2r^2}.
\end{equation}
It is possible to derive a closed-form solution for the evolution time
as a function of separation by starting with the integral formulation,
\begin{equation}
  t(r) = \int_{1}^{r} \frac{dr}{v(r)}.
\end{equation}
The velocity $v$ (in dimensionless units) can be found by noting that 
$\ddot{r} = v\, dv / dr$ and then separating and integrating the equation 
of motion. This yields 
\begin{equation}
  v(r) = \sqrt{\left(\frac{1}{r} - 1\right)}.
\end{equation}
For our problem $0 \leq r \leq 1$, so this is always valid. Integrating, we find
\begin{equation}
  t(r) = \text{arccos}\left(\sqrt{r}\right) + \sqrt{r \left(1 - r\right)}. \label{analyticalFreeFall}
\end{equation}
so that the point of contact would occur at $t = 1$. We actually stop the simulation
at $t = 0.9$, which is when the effects from the extended sizes of the stars
starts to become important. The results of our simulation for a $128^3$ zone 
uniform grid are shown in Figure \ref{Fig:Free Fall}. They show excellent agreement
between the theory and the simulation results, even for this very modest resolution.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{freefall/plot_freefall}
  \caption{Time evolution of two initially stationary white dwarfs,
    mutually attracted to each other by the gravitational force. The
    horizontal axis gives the separation of the white dwarfs, scaled
    to the initial separation, and the vertical axis gives the elapsed
    time of the simulation, scaled to the time it would take two point masses
    to collide. The solid curve shows the analytical result,
    calculated from Newtonian mechanics, and the circles show the
    samples from the time evolution with CASTRO.}
  \label{Fig:Free Fall}
\end{figure}

>>>>

A simple dynamical test to verify the gravitation physics in CASTRO is
the case of gravitational free fall. We place two stars on the grid 
in the manner of Section \ref{sec:initial_models}. The distance $a$ between 
them corresponds to a chosen orbital period $T$, consistent with the total
system mass $M$, but we disable the rotational source terms so that 
the stars start at rest in an inertial reference frame. 
Thus the stars will simply begin moving toward each other.
As long as the stars remain approximately spherical, the stars can be 
treated as point masses (this only seriously breaks down after the stars
have come into contact). In dimensionless units where $r \to r / a$ and 
$t \to 2\sqrt{2}\pi t / T$, the simple free fall equation of motion governing the
distance $r$ between their centers of mass takes the form:
\begin{equation}
  \ddot{r}(t) = - \frac{1}{2r^2}.
\end{equation}
It is possible to derive a closed-form solution for the evolution time
as a function of separation by starting with the integral formulation,
\begin{equation}
  t(r) = \int_{1}^{r} \frac{dr}{v(r)}.
\end{equation}
The velocity $v$ (in dimensionless units) can be found by noting that 
$\ddot{r} = v\, dv / dr$ and then separating and integrating the equation 
of motion. This yields 
\begin{equation}
  v(r) = \sqrt{\left(\frac{1}{r} - 1\right)}.
\end{equation}
For our problem $0 \leq r \leq 1$, so this is always valid. Integrating, we find
\begin{equation}
  t(r) = \text{arccos}\left(\sqrt{r}\right) + \sqrt{r \left(1 - r\right)}. \label{analyticalFreeFall}
\end{equation}
so that the point of contact would occur at $t = 1$. We actually stop the simulation
at $t = 0.9$, which is when the effects from the extended sizes of the stars
starts to become important. The results of our simulation for our default $256^3$ zone 
uniform grid are shown in Figure \ref{Fig:Free Fall}. They show excellent agreement
between the theory and the simulation results, and this agreement holds even for 
a more moderate $128^3$ zone simulation.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/freefall}
  \caption{Time evolution of two initially stationary white dwarfs,
    mutually attracted to each other by the gravitational force. The
    horizontal axis gives the separation of the white dwarfs, scaled
    to the initial separation, and the vertical axis gives the elapsed
    time of the simulation, scaled to the time it would take two point masses
    to collide. The solid curve shows the analytical result,
    calculated from Newtonian mechanics, and the circles show the
    samples from the time evolution with CASTRO. For visual clarity, we 
    show only a small fraction of the timesteps.}
  \label{Fig:Free Fall}
\end{figure}
